Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-11 paper/paper.tex
Timestamp and preallocation sections filled out.

\subsection{Timestamping stores}

The store delta technique removes overhead of joining large stores,
but for determining if a state has been visited before, we still have
to compare entire stores - an expensive computation.  We modify a
technique created in Shivers'
dissertation~\cite{ianjohnson:Shivers:1991:CFA}, timestamp
approximation, that removes the need for joining whole stores by
manipulating a global store imperatively. The original technique loses
precision because the starting store for each state in a machine
configuration can change between stepping all states and introduce
more flows than originally possible. However, by delaying the store
changes until after all steps have completed, we can maintain our
precision. In a variety of experiments, we observed that the overhead
for storing store-deltas was outweighed by the precision keeping flows
in check. That is to say, the analysis was faster {\it and} more precise.

TODO: technical detail

\subsection{Pre-allocating the store}
TODO

>>>>

\subsection{Timestamping an imperative store}

The store delta technique removes overhead of joining large stores,
but for determining if a state has been visited before (vis. $f(X)
\setminus S$ in ${\mathcal G}$), we still have to compare entire stores - an
expensive computation.  We modify a technique created in Shivers'
dissertation~\cite{ianjohnson:Shivers:1991:CFA}, timestamp
approximation, that removes the need for joining whole stores by
manipulating a global store imperatively. The original technique loses
precision because the starting store for each state in a machine
configuration can change between stepping all states and introduce
more flows than originally possible. However, by delaying the store
changes until after all steps have completed, we can maintain our
precision. In a variety of experiments, we observed that the overhead
for storing store-deltas was outweighed by the precision keeping flows
in check. That is to say, the analysis was faster {\it and} more
precise.

We pair this technique with a mutable workset. That is, to avoid the
intermediate sets introduced by the $State \to \Set(State)$ signature
of $\machstep$, we change the fixpoint algorithm to use a global
workset, {\tt todo}, that $\machstep$ imperatively updates with the
states that it reaches.

In total, we have six globals:
\begin{itemize}
 \item{{\tt $\msto$}, the store}
 \item{{\tt todo}, the workset}
 \item{{\tt seen}, a map from state to the timestamp at which it was last seen}
 \item{{\tt $\msdiff$}, the store changes for the current step}
 \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
 \item{{\tt T}, the timestamp of the store}
\end{itemize}

To ensure termination, we guard adding $c$ to {\tt todo} by the following check:
\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

After all steps complete, we apply $\msdiff$ to $\msto$ and increase
{\tt T} as long as there was at least one change in $\msdiff$. This
logic is only sound if we know that each $\ttuple{\maddr}{\mval{s}}$
in $\msdiff$ would change the value of $\maddr$ in the current
store. Thus, we also guard additions to $\msdiff$ so that only updates
that would change the store are permitted. Each time $\msdiff$ is
successfully extended, we set {\tt $\msdiff$?} to true. Before each
individual step, we set it to false.

\begin{figure}
\begin{alltt}
eval(\(e\)) :=
  \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
  \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \kmt\))
  while(true):
    if todo = \(\varnothing\): return (keys(seen), \(\msto\))
    else:
      let old := todo
      todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
      foreach c \(\in\) old: \(\msdiff\)? := false; c()
      unless \(\msdiff\) = \(\mtlst\): T += 1; \(\msdiff(\msto)\)
\end{alltt}
\caption{Imperative fixpoint algorithm}
\label{fig:imperative}
\end{figure}

\subsection{Pre-allocating the store}
Hashes are not the most efficent for raw speed. If we commit to using
sequential natural numbers for variables and expression labels, by the
time we finish parsing the input program, we will know exactly how
many addresses we will need. Thus the store can be represented by a
vector the size of the largest label assigned, plus 1. This technique
only works in 0-CFA, since there are many more addresses that can be
allocated in 1-CFA and higher. To support general $k$-CFA in a vector
representation, we need a growable vector and a map from addresses to
assigned indices. When assignable indices run out, we grow the
vector. The cost of this map is less than a hash representation for
the whole store since typically there are more references to an address
than allocations of it.
